Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-12-03 paper.qmd
Initial work on replacing the tSNE with UMAP

<!--UMAP applied for Two curvy data-->
```{r}
#| label: umap-two-curvy
umap_two_curvy <- read_rds(file = "data/two_non_linear_diff_shaped_close_clusters/two_non_linear_diff_shaped_close_clusters_umap_n-neigbors_15_min-dist_0.1.rds")

@@

nldr_two_curvy <- umap_two_curvy_scaled |> 
  ggplot(aes(x = UMAP1, y = UMAP2)) + 
  geom_point(alpha=0.5, colour="#636363", size = 0.5) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("a", sc_ltr_pos)

nldr_two_curvy2 <- umap_two_curvy_scaled |> 
  ggplot(aes(x = UMAP1, y = UMAP2)) + 
  geom_point(alpha=0.5, colour="#636363", size = 0.5) +
  interior_annotation("a", sc_ltr_pos)

@@

nldr_two_curvy_original2 <- umap_two_curvy_scaled |> 
  ggplot(aes(x = UMAP1, y = UMAP2)) + 
  geom_point(alpha=0.5, color = "#636363", size = 0.5) +
  theme(aspect.ratio = 1) +
  #xlim(c(-0.01, 1.1)) + ylim(c(-0.08, 0.57)) +
  interior_annotation("a1", sc_ltr_pos)
```

@@

hex_grid_two_curvy <- ggplot(
  data = hex_grid_with_counts, 
  aes(x = x, y = y)) +
  geom_polygon(color = "grey70", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = umap_two_curvy_scaled, 
             aes(x = UMAP1, y = UMAP2), 
             alpha = 0.5, size = 0.5, color = "#636363") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b", sc_ltr_pos)

hex_grid_coloured_two_curvy1 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts, 
    aes(x = x, y = y,
      group = hex_poly_id, 
      fill = std_counts), color = "grey70", linewidth=0.2) +
  geom_point(data = umap_two_curvy_scaled,
             aes(x = UMAP1, y = UMAP2),
             alpha = 0.3,
             size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "E") +
  interior_annotation("a", sc_ltr_pos) + 
  xlim(c(-0.3, 1.2)) + ylim(c(-0.17, 0.78))

@@

trimesh_two_curvy_removed1_umap_with_data <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = umap_two_curvy_scaled,
             aes(
               x = UMAP1,
               y = UMAP2
             ),
             color = "#636363",
             alpha = 0.5,
             size = 0.5) +
  #coord_equal() +
  # theme(
  #   aspect.ratio = 1
  #   ) +
  interior_annotation("b", sc_ltr_pos)
```

@@

hex_grid_coloured_two_curvy2 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_two_curvy2, 
    aes(x = x, y = y, 
        group = hex_poly_id, 
        fill = std_counts), 
        color = "grey70", 
        linewidth=0.2) +
  geom_point(data = umap_two_curvy_scaled,
           aes(x = UMAP1, y = UMAP2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "C") +
  xlim(c(-0.3, 1.2)) + ylim(c(-0.17, 0.78)) +
  interior_annotation("b", sc_ltr_pos)

@@

trimesh_two_curvy2 <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy2,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = umap_two_curvy_scaled,
             aes(
               x = UMAP1,
               y = UMAP2
             ),
             color = "#636363",
            alpha = 0.5,
            size = 0.5
            ) +
  coord_equal() +
  interior_annotation("a1", sc_ltr_pos)

@@

trimesh_two_curvy_removed1 <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy3,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000")+
  geom_point(data = umap_two_curvy_scaled,
             aes(
               x = UMAP1,
               y = UMAP2
             ),
             color = "#636363",
            alpha = 0.5,
            size = 0.5
            ) +
  coord_equal() +
  interior_annotation("c1", sc_ltr_pos)

trimesh_two_curvy_removed1_with_data <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy3,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = umap_two_curvy_scaled,
             aes(
               x = UMAP1,
               y = UMAP2
             ),
             color = "#636363",
            alpha = 0.5,
            size = 0.5
            ) +
  coord_equal() +
  #xlim(c(-0.01, 1.1)) + ylim(c(-0.08, 0.57)) +
  interior_annotation("a2", sc_ltr_pos)

@@

trimesh_two_curvy_removed2 <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy4,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000")+
  geom_point(data = umap_two_curvy_scaled,
             aes(
               x = UMAP1,
               y = UMAP2
             ),
             color = "#636363",
            alpha = 0.5,
            size = 0.5
            ) +
  coord_equal() +
  interior_annotation("b1", sc_ltr_pos)

@@

hex_grid_coloured_two_curvy3 <-  ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_s_curve3, 
    aes(x = x, y = y, 
        group = hex_poly_id, 
        fill = std_counts), 
        color = "grey70", linewidth=0.2) +
  geom_point(data = umap_two_curvy_scaled,
           aes(x = UMAP1, y = UMAP2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "D") +
  xlim(c(-0.3, 1.2)) + ylim(c(-0.17, 0.78)) +
  interior_annotation("c", sc_ltr_pos)
```

@@

error_plot_two_curvy <- error_df_two_curvy_abs |>
  ggplot(aes(x = UMAP1,
             y = UMAP2,
             colour = sqrt_row_wise_abs_error)) +
  geom_point(alpha=0.5) +
  scale_colour_continuous_sequential(palette = "YlOrRd", n_interp = 20) +
  theme(
    aspect.ratio = 1
  )

@@

```{r}
#| echo: false
#| label: fig-predict-two-curvy
#| fig-pos: H
#| fig-width: 10
#| fig-height: 10
#| out-height: 80%
#| fig-cap: "Comparison of prediction generated using the exiting `umap` R package's prediction method and our method: (a) A view of the true model in projection from $7\\text{-}D$, (b) predicted data from the `umap` prediction method, and (c) predicted data from our method."

>>>>

<!--UMAP applied for Two curvy data-->
```{r}
#| label: umap-two-curvy
umap_two_curvy <- read_rds(file = "data/two_non_linear_diff_shaped_close_clusters/two_non_linear_diff_shaped_close_clusters_tsne_perplexity_30.rds")

@@

nldr_two_curvy <- umap_two_curvy_scaled |> 
  ggplot(aes(x = tSNE1, y = tSNE2)) + 
  geom_point(alpha=0.5, colour="#636363", size = 0.5) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("a", sc_ltr_pos)

nldr_two_curvy2 <- umap_two_curvy_scaled |> 
  ggplot(aes(x = tSNE1, y = tSNE2)) + 
  geom_point(alpha=0.5, colour="#636363", size = 0.5) +
  interior_annotation("a", sc_ltr_pos)

@@

nldr_two_curvy_original2 <- umap_two_curvy_scaled |> 
  ggplot(aes(x = tSNE1, y = tSNE2)) + 
  geom_point(alpha=0.5, color = "#636363", size = 0.5) +
  theme(aspect.ratio = 1) +
  #xlim(c(-0.01, 1.1)) + ylim(c(-0.08, 0.57)) +
  interior_annotation("a1", sc_ltr_pos)
```

@@

hex_grid_two_curvy <- ggplot(
  data = hex_grid_with_counts, 
  aes(x = x, y = y)) +
  geom_polygon(color = "grey70", 
               aes(group = hex_poly_id), 
               fill = "#ffffff") +
  geom_point(data = umap_two_curvy_scaled, 
             aes(x = tSNE1, y = tSNE2), 
             alpha = 0.5, size = 0.5, color = "#636363") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b", sc_ltr_pos)

hex_grid_coloured_two_curvy1 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts, 
    aes(x = x, y = y,
      group = hex_poly_id, 
      fill = std_counts), color = "grey70", linewidth=0.2) +
  geom_point(data = umap_two_curvy_scaled,
             aes(x = tSNE1, y = tSNE2),
             alpha = 0.3,
             size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "E") +
  interior_annotation("a", sc_ltr_pos) + 
  xlim(c(-0.3, 1.2)) + ylim(c(-0.17, 0.78))

@@

trimesh_two_curvy_removed1_umap_with_data <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = umap_two_curvy_scaled,
             aes(
               x = tSNE1,
               y = tSNE2
             ),
             color = "#636363",
             alpha = 0.5,
             size = 0.5) +
  #coord_equal() +
  # theme(
  #   aspect.ratio = 1
  #   ) +
  interior_annotation("b", sc_ltr_pos)
```

@@

hex_grid_coloured_two_curvy2 <- ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_two_curvy2, 
    aes(x = x, y = y, 
        group = hex_poly_id, 
        fill = std_counts), 
        color = "grey70", 
        linewidth=0.2) +
  geom_point(data = umap_two_curvy_scaled,
           aes(x = tSNE1, y = tSNE2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, 
    na.value = "#ffffff", option = "C") +
  xlim(c(-0.3, 1.2)) + ylim(c(-0.17, 0.78)) +
  interior_annotation("b", sc_ltr_pos)

@@

trimesh_two_curvy2 <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy2,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = umap_two_curvy_scaled,
             aes(
               x = tSNE1,
               y = tSNE2
             ),
             color = "#636363",
            alpha = 0.5,
            size = 0.5
            ) +
  coord_equal() +
  interior_annotation("a1", sc_ltr_pos)

@@

trimesh_two_curvy_removed1 <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy3,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000")+
  geom_point(data = umap_two_curvy_scaled,
             aes(
               x = tSNE1,
               y = tSNE2
             ),
             color = "#636363",
            alpha = 0.5,
            size = 0.5
            ) +
  coord_equal() +
  interior_annotation("c1", sc_ltr_pos)

trimesh_two_curvy_removed1_with_data <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy3,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000") +
  geom_point(data = umap_two_curvy_scaled,
             aes(
               x = tSNE1,
               y = tSNE2
             ),
             color = "#636363",
            alpha = 0.5,
            size = 0.5
            ) +
  coord_equal() +
  #xlim(c(-0.01, 1.1)) + ylim(c(-0.08, 0.57)) +
  interior_annotation("a2", sc_ltr_pos)

@@

trimesh_two_curvy_removed2 <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy4,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000")+
  geom_point(data = umap_two_curvy_scaled,
             aes(
               x = tSNE1,
               y = tSNE2
             ),
             color = "#636363",
            alpha = 0.5,
            size = 0.5
            ) +
  coord_equal() +
  interior_annotation("b1", sc_ltr_pos)

@@

hex_grid_coloured_two_curvy3 <-  ggplot() + 
  geom_polygon(
    data = hex_grid_with_counts_s_curve3, 
    aes(x = x, y = y, 
        group = hex_poly_id, 
        fill = std_counts), 
        color = "grey70", linewidth=0.2) +
  geom_point(data = umap_two_curvy_scaled,
           aes(x = tSNE1, y = tSNE2),
           alpha = 0.3,
           size = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "D") +
  xlim(c(-0.3, 1.2)) + ylim(c(-0.17, 0.78)) +
  interior_annotation("c", sc_ltr_pos)
```

@@

error_plot_two_curvy <- error_df_two_curvy_abs |>
  ggplot(aes(x = tSNE1,
             y = tSNE2,
             colour = sqrt_row_wise_abs_error)) +
  geom_point(alpha=0.5) +
  scale_colour_continuous_sequential(palette = "YlOrRd", n_interp = 20) +
  theme(
    aspect.ratio = 1
  )

@@

```{r}
#| echo: false
#| label: fig-predict-two-curvy
#| fig-pos: H
#| fig-width: 10
#| fig-height: 10
#| out-height: 80%
#| fig-cap: "Comparison of prediction generated using the exiting `umap` R package's prediction method and our method: (a) A view of the true model in projection from $7\\text{-}D$, (b) predicted data from the `umap` prediction method, and (c) predicted data from our method."
#| eval: false
